---
title: "Final Project"
author: "Julio A Canas"
date: "12/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Reference Guide {.tabset .tabset-pills}


## Loading Data and Packages

### here( ), library and require

The here package enables easy file referencing by using the top-level directory of a file project to easily build file paths. here() uses a reasonable heuristics to find your project's files, based on the current working directory at the time when the package is loaded. Also, to load and attach add-on packages "library" and "require" are needed. 

library(package) and require(package) both load the namespace of the package with name package and attach it on the search list. require is designed for use inside other functions; it returns FALSE and gives a warning (rather than an error as library() does by default) if the package does not exist. Both functions check and update the list of currently attached packages and do not reload a namespace which is already loaded.

An example of how to load packages are here and palmerpenguins, note that you load them different:
```{r}
#here will display your top-level directory
library("here")
require("here")

#palmerpenguins will load the palmerpenguins package
library(palmerpenguins)
require(palmerpenguins)

```

### read.csv( )

This code reads a file in table format and creates a data frame from it, with cases corresponding to lines and variables to fields in the file. Most of the time, you will also need the previous information of here(), library and require in order to load the correct document.

```{r}
library("here")
require("here")
here()

#To load the information in the csv to the data frame named ginko, from the top-level directory
ginkgo=read.csv(here("data","ginkgo_data_2021.csv"))

#To obtain information of the data frame we use head() and summary()
head(ginkgo)
summary(ginkgo)
```

## Data Structures

### c( )
The function c() combines or concatenates its arguments into a vector (a 1-dimensional data structure consisting of 1 or more elements).

* All of the elements must be of the same type.
  * I can’t combine character and numeric types in the same call to c()

Here’s two examples using numeric and character data types:
```{r}
## Create a vector of numbers:
num_vec  = c(1, 4, 8, 9, 13)

## Create a vector of characters:
char_vec = c("a", "fish", "data is cool")
```

I can show the contents of a vector by typing the name of the vector, or using the print() function.
```{r}
## Typing the name of the vector into the console prints the contents
num_vec
```

```{r}
## The print() function accomplishes the same task:
print(char_vec)
```

### length( )

Get or set the length of vectors (including lists) and factors, and of any other R object for which a method has been defined.

Let's use the vectors of the c() section:
```{r}
#Remember that num_vec  = c(1, 4, 8, 9, 13) is a 5 element vector
length(num_vec)

#Remember that char_vec = c("a", "fish", "data is cool") is a 3 element vector
length(char_vec)
```

### matrix( )

A matrix is a two dimensional data structure in R programming. 

Matrix can be created using the matrix() function. Dimension of the matrix can be defined by passing appropriate value for arguments nrow (number of rows) and ncol (number of columns).

Providing value for both dimension is not necessary. If one of the dimension is provided, the other is inferred from length of the data. Also, we can see that the matrix is filled column-wise, this can be reversed to row-wise filling by passing TRUE to the argument byrow.

```{r}
#Different value arragements in a matrix with the same values and typing
matrix(1:12, nrow=4, ncol=3, byrow=TRUE) #or matrix(1:12, nrow=4, byrow=TRUE)

matrix(1:12, nrow=4, ncol=3, byrow=FALSE) #or matrix(1:12, nrow=4, byrow=FALSE)

matrix(1:12, nrow=3, ncol=4, byrow=TRUE) #or matrix(1:12, nrow=3, , byrow=TRUE)

matrix(1:12, nrow=3, ncol=4, byrow=FALSE) #or matrix(1:12, nrow=3, byrow=FALSE)
```

Finally, you can also create a matrix from a vector by setting its dimension using dim(). The first number in the dim() output is the number of rows, and the second number of the number of columns.
```{r}
x<-c(1,2,3,4,5,6,7,8,9,10)
dim(x)<-c(2,5)
x
```


### data.frame( )

The function data.frame() creates data frames, tightly coupled collections of variables which share many of the properties of matrices and of lists, used as the fundamental data structure by most of R's modeling software.

```{r}
#You can use a vector from the c ( ) section and add more information (as long as there are vectors with the same amount of elements as the one used or you will have an error)
my.frame<-data.frame(num_vec, abc_vec=c("A", "B", "C", "D", "E"))

my.frame #to print the data.frame
```

### Examples

```{r}
#The nrow() function returns the number of rows in your object.
nrow(ginkgo)

# The ncol() function returns the number of columns in your object.
ncol(ginkgo)

#The dim( ) function can  create a matrix from a vector by setting its dimension 
dim(ginkgo)
```

## Subsetting

### $

The $ operator is used to extract or subset a specific part of a data object in R. For instance, this can be a data frame object or a list.

```{r}
#To get the subset of the first column of the ginkgo data we employ this code
subset_ginkgo_site_id<-ginkgo$site_id
subset_ginkgo_site_id
```

### [ ]

Using the ‘[ ]’ operator, elements of vectors and observations from data frames can be accessed. To neglect some indexes, ‘-‘ is used to access all other indexes of vector or data frame. 

```{r}
#To show the first row we use [1,], this includes every column
ginkgo[1,]

#To show the element in the second row and third column we use
ginkgo[3,2]

#To show the third column we use [,3], this includes every value of the 756 rows in total
ginkgo[,3]
```

### subset( )

When looking to create more complex subsets or a subset based on a condition, the next step up is to use the subset() function.

```{r}
#First we need to upload the palmerpenguins in R
require(palmerpenguins)

#To etrieve all the data for Adelie penguins (in the species column) from the peuguins dataset we use
dat_adelie = droplevels(subset(penguins, species == "Adelie"))
dat_adelie
```

## Numerical Data Exploration

### summary( )

The summary() function in R can be used to quickly summarize the values in a vector, data frame, regression model, or ANOVA model in R.

```{r}
#Let's use summary() with the ginkgo data frame and da_adelie from palmerpenguins csv
summary(ginkgo)
summary(dat_adelie)
```

### mean( )

The function mean() is used to calculate the mean of a set by taking the sum of the values and dividing with the number of values in a data series.

```{r}
#Let's use mean( ) with the subset of max_width ginkgo data frame (the function only works with numerical data)
mean(ginkgo$max_width)
```

### sd( )

Standard deviation in R is a statistic that measures the amount of dispersion or variation of a set of value, generally, it is used when we are dealing with values where we have to find the difference between the values and the mean.

```{r}
#Let's use mean( ) with the subset of max_width ginkgo data frame (the function only works with numerical data)
sd(ginkgo$max_width)
```

## Graphical Data Exploration

### scatterplot( )

Scatterplot: Using the ginkgo data, reate a scatterplot of max leaf depth (x) and max leaf width (y).
You can personalize your plot with the following commands:

* col: defines color of the plots in the scatterplot.
* pch: defines the point symbols in the functions plot() and lines(). The pch stands for plot character.  The pch contains numeric values rangers from 0 to 25 or character symbols (“+”, “.”, “;”, etc.) specifying in symbols (or shapes).
* cex: defines a number indicating the amount by which plotting text and symbols should be scaled relative to the default.
* main: defines the title of the plot
* xlab: defines the name of the x axis in the plot
* ylab: defines the name of the y axis in the plot
* xlim: defines the limits of the scale in the x axis
* ylim: defines the limits of the scale in the y axis

```{r}
#Let's plot the maximum depth and width of the ginkgo data frame
plot(ginkgo$max_depth, ginkgo$max_width,
     col="red",
     pch=0, #2 stands for squares
     cex=1, #we won't increase the size for better observation of the plots
     main="Ginkgo Max depth vs Max width",
     xlab="Ginkgo max_depth",
     ylab="Ginkgo max_width",
     # xlimit=c(0,160), (we won't include it because of the warnings to be displayed)
     # ylimit=c(0,140) (we won't include it because of the warnings to be displayed)
     )
```

### hist( )

The generic function hist computes a histogram of the given data values. If plot = TRUE, the resulting object of class "histogram" is plotted by plot.histogram, before it is returned.

You can personalize your plot with the following commands:
* breaks: a single number giving the number of cells for the histogram.

```{r}
#Let's make a histogram of the penguin Adelie flipper lengths from the palmerpenguins data
hist(dat_adelie$flipper_length_mm,
     breaks=8,
     main="Histogram of Adelie's flipper lengths",
     xlab="Adelie flipper lenght (mm)")
```

### boxplot( )

In R, boxplot  (and whisker plot) is created using the boxplot() function. The boxplot() function takes in any number of numeric vectors, drawing a boxplot for each vector.

```{r}
#To create a boxplot we can use:
boxplot(ginkgo$petiole_length,
        main="Boxplot of ginkgos's petiole lenght",
        ylab="Petiole lenght (mm)")

#To create a conditional boxplot we can use:
boxplot(ginkgo$petiole_length~ginkgo$seeds_present,
        main="Boxplot of ginkgos's petiole length with seed presence",
        xlab="Seeds present (F or T)",
        ylab="etiole length (mm)")
```

### Par( )

You can use the par() function in R to create multiple plots at once.

```{r}
#Now we can plot every graph we had made so far in a single plot
par(mfrow=c(2,2))
plot(ginkgo$max_depth, ginkgo$max_width,
     col="red",
     pch=0, #2 stands for squares
     cex=1, #we won't increase the size for better observation of the plots
     main="Ginkgo Max depth vs Max width",
     xlab="Ginkgo max_depth",
     ylab="Ginkgo max_width",
     # xlimit=c(0,160), (we won't include it because of the warnings to be displayed)
     # ylimit=c(0,140) (we won't include it because of the warnings to be displayed)
     )
hist(dat_adelie$flipper_length_mm,
     breaks=8,
     main="Histogram of Adelie's flipper lengths",
     xlab="Adelie flipper lenght (mm)")
boxplot(ginkgo$petiole_length,
        main="Boxplot of ginkgos's petiole lenght",
        ylab="Petiole lenght (mm)")
boxplot(ginkgo$petiole_length~ginkgo$seeds_present,
        main="Boxplot of ginkgos's petiole length with seed presence",
        xlab="Seeds present (F or T)",
        ylab="etiole length (mm)")
```

## Distribution function

### dnorm( )

The function dnorm returns the value of the probability density function (pdf) of the normal distribution given a certain random variable x, a population mean μ and population standard deviation σ. The syntax for using dnorm is as follows: dnorm(x, mean, sd) 

```{r}
#Calculate the probability of a normal distribution with x=5, mean=20 and sd=2
dnorm(x=5, mean=20, sd=2)
```

### pnorm( )

The function pnorm returns the value of the cumulative density function (cdf) of the normal distribution given a certain random variable q, a population mean μ and population standard deviation σ. The syntax for using pnorm is as follows: pnorm(q, mean, sd) 

```{r}
#Calculate the percentage of females heavier than 50 pounds with a mean of 45 and a sd of 5
pnorm(50, mean=42, sd=5)
```

### qnorm( )

The function qnorm returns the value of the inverse cumulative density function (cdf) of the normal distribution given a certain random variable p, a population mean μ and population standard deviation σ. The syntax for using qnorm is as follows: qnorm(p, mean, sd)

```{r}
#Find the Z-score of the 89th quantile of the standard normal distribution 
qnorm(0.89, mean=0, sd=1)
```

### dbinom( )

The function dbinom returns the value of the probability density function (pdf) of the binomial distribution given a certain random variable x, number of trials (size) and probability of success on each trial (prob). The syntax for using dbinom is as follows: dbinom(x, size, prob) 

```{r}
# Calculate the probability of a 15 success during 20 trials where the probability of success is 0.64
dbinom(x=15, size=20, prob=0.64)
```

### pbinom( )

The function pbinom returns the value of the cumulative density function (cdf) of the binomial distribution given a certain random variable q, number of trials (size) and probability of success on each trial (prob). The syntax for using pbinom is as follows: pbinom(q, size, prob) 

```{r}
#Calculate the probability of 5 or fewer success during a 9 trials where the probability of success is 0.47
pbinom(5, size=9, prob=0.47)
```

### qbinom( )

The function qbinom returns the value of the inverse cumulative density function (cdf) of the binomial distribution given a certain random variable q, number of trials (size) and probability of success on each trial (prob). The syntax for using qbinom is as follows: qbinom(q, size, prob) 

```{r}
#Find the 58th quartile of a binomial distribution with 45 trials and probability of success of 0.3
qbinom(0.58, size=45, prob=0.3)
```

```